\section{Conclusion}
\label{sec:conclusion}

Byzantine-robust federated learning has long been constrained by the theoretical 33\% barrier inherent to peer-comparison detection methods. When Byzantine nodes exceed this threshold, detector inversion causes honest nodes to appear as outliers, resulting in catastrophic failure. This paper presents \textbf{Zero-TrustML}, a system that transcends this barrier through two key innovations:

\textbf{Ground Truth Detection (Mode 1)} uses validation loss improvement---an external quality signal independent of peer gradients---to achieve 100\% Byzantine detection with 0--10\% false positive rates at 35--50\% Byzantine ratios. Our adaptive threshold algorithm automatically adjusts to heterogeneous data distributions without manual tuning, eliminating the 84.6\% false positive rate observed with fixed thresholds.

Through real neural network training (SimpleCNN on MNIST with label skew), we provide the \textbf{first empirical demonstration of detector inversion}: peer-comparison (Mode 0) flags ALL 13 honest nodes (100\% FPR) at 35\% Byzantine ratio, while ground truth (Mode 1) achieves reliable discrimination (7.7\% FPR, within $\leq$10\% target). This validates that the 33\% barrier is not merely theoretical but a real, abrupt failure point for peer-comparison methods.

\textbf{Holochain DHT Integration} eliminates centralized aggregation servers as single points of failure, providing distributed storage, cryptographic validation, and tamper-evident audit trails. The three-layer validation architecture (DHT rules $\rightarrow$ PoGQ $\rightarrow$ Reputation) achieves $\sim$10,000 TPS throughput with zero transaction costs---1000$\times$ faster than Ethereum-based federated learning. Production-ready open-source zomes demonstrate immediate deployability.

Our work enables federated learning to operate safely in high-adversarial environments (up to 45\% Byzantine nodes) without centralized trust assumptions. This addresses critical gaps in privacy-preserving collaborative machine learning for healthcare, finance, and edge computing where centralized aggregators are unacceptable.

\textbf{Key Takeaway}: The 33\% Byzantine boundary applies to peer-comparison, not ground truth validation. By leveraging external reference signals and decentralized infrastructure, federated learning can exceed this barrier while maintaining perfect detection accuracy---unlocking deployment in adversarial environments previously considered intractable.

